---
title: "Course Project 1"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

# RepData - Peer Assignment 1

## Overview
This analysis looks at the number of steps per day from wearable devices - see README.MD in the github repo for more information  

## Load libraries

Libraries needed for analysis  

```{r libraries, echo=TRUE}
library(plyr)
library(Hmisc)
library(lattice)
setwd("C:/Users/Alex/Documents/KNITR")
```

## Read in CSV data

Data downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  
Data unzipped and loaded from directory below  

```{r readdata, echo=TRUE}
dataActivity <- read.csv("activity.csv", header=TRUE)
```

## Histogram of Number of Steps per Day

Work out sum of steps for each date - plot on histogram  

```{r stepsperday, echo=TRUE}
stepsPerDay <- aggregate(steps~date,data=dataActivity, sum)
hist(x=stepsPerDay$steps, xlab="Steps per Day", main="Histogram of Steps per Day")
```

## Calculation of Mean and Median Steps

Calculate the mean, median and total steps per day  

```{r meanmedian, echo=TRUE}
meanSteps <- mean(stepsPerDay$steps)
medianSteps <- median(stepsPerDay$steps)
totalSteps <- sum(stepsPerDay$steps)
```
The mean steps per day is `r meanSteps`  
The median steps per day is `r medianSteps`  
The total number of steps per day is `r totalSteps`  

## Plot of Steps per Interval  

Calculate and plot the average steps per interval  

```{r stepsperinterval, echo=TRUE}
stepsPerInterval <- aggregate(steps~interval,data=dataActivity, FUN=mean)
plot(steps ~ interval, xlab="Interval", ylab="Steps", main="Average Steps per Interval", data=stepsPerInterval, type="l")
```

## Interval that contains the maximum number of steps  

Find the interval with the maximum number of steps  

```{r intervalmaxsteps, echo=TRUE}
intervalMaxSteps <- stepsPerInterval[stepsPerInterval$steps >= max(stepsPerInterval$steps), 1]
```
The interval with the max steps is `r intervalMaxSteps`

## Number of NAs  

Calculate the number of records with NA in the data  

```{r numberofNAs, echo=TRUE}
numberOfNAs <- sum(is.na(dataActivity$steps))
```
The number of NAs in the data is `r numberOfNAs`

## Impute on NAs - Create new Imputed data set for NAs

Save the mean when there is an NA - use impute function to do this

```{r ImputeonNAs, echo=TRUE}
dataActivityImputed <- dataActivity
dataActivityImputed$stepsImputed <- with(dataActivity,impute(dataActivity$steps, mean))
dataActivityImputed$steps <- dataActivityImputed$stepsImputed
```

## Make new histogram off Imputed data set

calculate and plot the __imputed__ steps per day

```{r stepsperdayimpute, echo=TRUE}
stepsPerDayImputed <- aggregate(steps~date,data=dataActivityImputed, sum)
hist(x=stepsPerDayImputed$steps, xlab="Steps per Day", main="Histogram of Steps per Day, with Imputed Values")
```  
Note that this looks very close to previous plot with NAs

## Mean and Median on Imputed data set - Comparison

Calculate the mean, median and total steps

```{r meanmedianimputed, echo=TRUE}
meanStepsImputed <- mean(stepsPerDayImputed$steps)
medianStepsImputed <- median(stepsPerDayImputed$steps)
totalStepsImputed <- sum(stepsPerDayImputed$steps)
```  

## Comparing the results  
__Mean data the same__  
Mean Imputed Data `r meanStepsImputed`, Mean Non-imputed Data `r meanSteps`  
__Median data increased slightly__  
Median Imputed Data `r medianStepsImputed`, Median Non-imputed Data `r medianSteps`  
__Total steps increased significantly__  
Total Steps Imputed Data `r totalStepsImputed`, Total Steps Non-imputed Data `r totalSteps`  


# Add Weekday and Weekend Indicators

Add weekday, weekend indicators

```{r weekdayindicator, echo=TRUE}
dataActivityImputed$Weekday <- weekdays(as.Date(dataActivityImputed$date))
dataActivityImputed$IsWeekend <- grepl('Saturday|Sunday', dataActivityImputed$Weekday)
dataActivityImputed$IsWeekday <- !grepl('Saturday|Sunday', dataActivityImputed$Weekday)
dataActivityImputed$WeekdayStatus <- ifelse(dataActivityImputed$IsWeekday==TRUE,"weekday","weekend")
```

# Panel plot - Weekday Steps vs Weekend Steps
Panel plot the average number of steps per interval, splitting by whether it is a weekend or a weekday  

```{r weekdaysteps, echo=TRUE}
stepsPerIntervalImputed <- aggregate(steps ~ interval + WeekdayStatus,data=dataActivityImputed, FUN=sum)
xyplot(steps~interval|factor(WeekdayStatus),type="l",layout=c(1,2), xlab="Interval",ylab="Number of Steps", data=stepsPerIntervalImputed)
```
